---
title: "Practical Machine Learning-Prediction Assignment"
author: "AM"
date: "12/12/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

##Installing Packages 
```{r}
#install.packages("caret")
#install.packages("randomForest")
#install.packages("rpart")
library(lattice)
library(ggplot2)
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
set.seed(1234)
```
##Loading data and cleaning
```{r}
# After saving both data sets into my working directory
# Some missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.
# We notice that both data sets contain columns with all missing values - these will be deleted.  

# Loading the training data set into my R session replacing all missing with "NA"
trainingset <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!", ""))

# Loading the testing data set 
testingset <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))

# Check dimensions for number of variables and number of observations
dim(trainingset)
dim(testingset)

# Delete columns with all missing values
trainingset<-trainingset[,colSums(is.na(trainingset)) == 0]
testingset <-testingset[,colSums(is.na(testingset)) == 0]

# Some variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). We can delete these variables.
trainingset<-trainingset[,-c(1:7)]
testingset<-testingset[,-c(1:7)]
# and have a look at our new datasets:
dim(trainingset)
dim(testingset)
head(trainingset)
head(testingset)
```
##Partitioning the training data set to allow cross-validation

The training data set contains 53 variables and 19622 obs.
The testing data set contains 53 variables and 20 obs.
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (75%) and subTest (25%).
This will be performed using random subsampling without replacement.

```{r}
subsamples <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
subTraining <- trainingset[subsamples, ] 
subTesting <- trainingset[-subsamples, ]
dim(subTraining)
dim(subTesting)
head(subTraining)
head(subTesting)
```

##A look at the data
The variable “classe” contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.
```{r}
plot(subTraining$classe, col="purple", main="Bar Plot of levels of the variable classe within the subTraining data set", xlab="classe levels", ylab="Frequency")
```
From the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent with more than 4000 occurrences while level D is the least frequent with about 2500 occurrences.

##First prediction model: Using Decision Tree
```{r}
model1<-rpart(classe ~ ., data=subTraining, method="class")

# Predicting:
prediction1 <- predict(model1, subTesting, type = "class")

# Plot of the Decision Tree
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

```{r}
# Test results on our subTesting data set:
confusionMatrix(prediction1, subTesting$classe)
```

##Second prediction model: Using Random Forest
```{r}
model2 <- randomForest(classe ~. , data=subTraining, method="class")

# Predicting:
prediction2 <- predict(model2, subTesting, type = "class")

# Test results on subTesting data set:
confusionMatrix(prediction2, subTesting$classe)
```

##Decision

As expected, Random Forest algorithm performed better than Decision Trees.

```{r}
# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(model2, testingset, type="class")
predictfinal
```


